Estimation of Spatiotemporal Integro-Difference Equations for Dynamical Neural Field Models

### 2009-11-17 ltx/BrainIDE.tex
Added alternative state-space derivation with Heaviside function

The neural field models use statistics of the cortex to relate mean firing rates of pre-synaptic neural populations to post-synaptic mean membrane potentials. Each neural population represents a functional cortical processing unit referred to as a column. The columnar organization of the cortex is not discrete, but is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modeled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range corticocortical and corticothalamic connectivity)~\cite{Jirsa2009}~\cite{Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition~\cite{Braitenberg1998}. For example, this structural organization is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}.

@@

\begin{equation}\label{new_spatiotemporal_kernels}
\frac{{dv\left( {r,t} \right)}}{{dt}} = \int_\Omega  {k\left( {r - r'} \right)\tilde v\left( {r',t} \right) - \alpha \delta \left( {r - r'} \right)\bar v\left( {r,t} \right)dr'}  + e\left( {r,t} \right)
\end{equation}

@@

@@

 v_{t+1}(\mathbf s)=\int_{\mathcal S} \alpha \delta(\mathbf s-\mathbf r)\bar v_t(\mathbf r)d \mathbf r+\int_{\mathcal S} W(\mathbf s -\mathbf r)\tilde v_t(\mathbf r)d\mathbf r+e_t(\mathbf s)

@@

\bar v_t(\mathbf s)=\boldsymbol \phi^{\top}(\mathbf s)\bar{\mathbf x}_t

@@

\tilde v_t(\mathbf s)=\boldsymbol \phi^{\top}(\mathbf s)\tilde{\mathbf x}_t

@@

 W(\mathbf s -\mathbf r)=\boldsymbol \theta^{\top}\psi(\mathbf s -\mathbf r)

@@

@@

\subsection{Linear Model}

\subsection{Nonlinear Model}

>>>>

Neural field models use statistics of the cortex to relate mean firing rates of pre-synaptic neural populations to post-synaptic mean membrane potentials. Each neural population represents a functional cortical processing unit referred to as a column. The columnar organization of the cortex is not discrete, but is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modeled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range corticocortical and corticothalamic connectivity)~\cite{Jirsa2009}~\cite{Qubbaj2007}. Locally, each column is thought be connected via symmetric short range local excitation, with surround inhibition~\cite{Braitenberg1998}. For example, this structural organization is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}.

@@

For the state-space derivation, to create an estimator, the model can be reformulated like so
\begin{equation}\label{new_spatiotemporal_kernels}
\frac{{dv\left( {r,t} \right)}}{{dt}} = \int_\Omega  {k\left( {r - r'} \right)\tilde v\left( {r',t} \right) - \alpha \delta \left( {r - r'} \right)\bar v\left( {r,t} \right)dr'}  + e\left( {r,t} \right)
\end{equation}

@@

\begin{figure*}
\centering
\includegraphics[scale=0.4]{ExampleFieldDecomposition.eps}
\caption[ExampleFieldDecomposition]{Illustration of a one dimensional field decomposition using 5 Gaussian basis functions, each weighted an element of the state vector. Each basis function has a width of 0.01 and is uniformly positioned with a separation of 0.25 (with arbitrary units).}
\label{ExampleFieldDecomposition}
\end{figure*}

@@

 v_{t+1}(\mathbf s)=\int_{\mathcal S} \alpha \delta(\mathbf s-\mathbf r)\bar v_t(\mathbf r)d \mathbf r+\int_{\mathcal S} w(\mathbf s -\mathbf r)\tilde v_t(\mathbf r)d\mathbf r+e_t(\mathbf s)

@@

\bar v_t(\mathbf r)=\boldsymbol \phi^{\top}(\mathbf r)\bar{\mathbf x}_t

@@

\tilde v_t(\mathbf r)=\boldsymbol \phi^{\top}(\mathbf r)\tilde{\mathbf x}_t=\nu_0

@@

 w(\mathbf r -\mathbf r')=\boldsymbol \theta^{\top}\psi(\mathbf r -\mathbf r')

@@

\subsection{State-Space Model Alternative Approach}
First we define
\begin{equation}\label{DefineKappa}
\kappa \left( r' \right) = \left\{ {\begin{array}{*{20}{c}}
   {{\nu _0}} & {v\left( {r',t} \right) \ge {v_0}}  \\
   0 & {otherwise}  \\
\end{array}} \right.
\end{equation}

\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  {v_{t + 1}}\left( r \right)  &=& \int_\Omega  {k\left( {r - r'} \right)\kappa \left( r' \right) - \alpha \delta \left( {r - r'} \right)v\left( {r,t} \right)dr'}  + \varepsilon\left( {r,t} \right) \\
   &=& \int_\Omega  {k\left( {r - r'} \right)\kappa \left( r' \right)dr'}  - \alpha\int_\Omega  { \delta \left( {r - r'} \right)v\left( {r,t} \right)dr'}  + \varepsilon\left( {r,t} \right)
\end{eqnarray}
Defining
\begin{equation}\label{FieldDecomp}
{v_t}\left( r \right) = {\phi ^T}\left( r \right){x_t}
\end{equation}
\begin{equation}\label{KernelDecomp}
k\left( {r - r'} \right) = {\theta ^T}\psi \left( {r - r'} \right)
\end{equation}
Making substitutions
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  {\phi ^T}\left( r \right){x_{t+1}}  &=& {\theta ^T}\int_\Omega  {\psi \left( {r - r'} \right)\kappa \left( r' \right)dr'}  - \alpha \int_\Omega  {\delta \left( {r - r'} \right){\phi ^T}\left( r \right){x_t}dr'}  + \varepsilon \left( {r,t} \right) \\
   &=& {\theta ^T}\int_\Omega  {\psi \left( {r - r'} \right)\kappa \left( r' \right)dr'}  - \alpha {\phi ^T}\left( r \right){x_t} + \varepsilon \left( {r,t} \right)
\end{eqnarray}
Now we define
\begin{equation}\label{DefFiringConnectivity}
\Phi \left( r \right) = \int_\Omega  {\psi \left( {r - r'} \right)\kappa \left( r' \right)dr'}.
\end{equation}
Substituting in
\begin{equation}\label{Dunno}
{\phi ^T}\left( r \right){x_{t+1}} = {\theta ^T}\Phi \left( r \right) - \alpha {\phi ^T}\left( r \right){x_t} + \varepsilon \left( {r,t} \right)
\end{equation}
Cross multiply by $\phi(r)$ and integrating over $r$ we get
\begin{equation}\label{TooTired}
\int_\Omega  {\phi \left( r \right){\phi ^T}\left( r \right)dr} {x_{t+1}} = \int_\Omega  {\phi \left( r \right){\theta ^T}\Phi \left( r \right)dr}  - \alpha \int_\Omega  {\phi \left( r \right){\phi ^T}\left( r \right)dr} {x_t} + \int_\Omega  {\phi \left( r \right)\varepsilon \left( {r,t} \right)dr}
\end{equation}
Now let us define
\begin{equation}\label{DefinePsix}
{\Psi _x} = \int_\Omega  {\phi \left( r \right){\phi ^T}\left( r \right)dr}
\end{equation}
\begin{equation}\label{DefinePsiTheta}
{\Psi _\theta } = \int_\Omega  {\phi \left( r \right){\theta ^T}\Phi \left( r \right)dr}.
\end{equation}
Making substitutions
\begin{equation}\label{poo}
{\Psi _x}{x_{t+1}} = {\Psi _\theta } - \alpha {\Psi _x}{x_t} + \int_\Omega  {\phi \left( r \right)\varepsilon \left( {r,t} \right)dr}.
\end{equation}
Isolating the state we get
\begin{equation}\label{IsolatingState}
{x_{t+1}} = {\Psi _x}^{ - 1}{\Psi _\theta } - \alpha {x_t} + {\Psi _x}^{ - 1}\int_\Omega  {\phi \left( r \right)\varepsilon \left( {r,t} \right)dr}
\end{equation}

\begin{equation}\label{SSmodel}
{x_{t+1}} = A{x_t} + Bu + {\Psi _x}^{ - 1}\int_\Omega  {\phi \left( r \right)\varepsilon \left( {r,t} \right)dr}
\end{equation}
\begin{eqnarray}
% \nonumber to remove numbering (before each equation)
  A &=& -\alpha \\
  B &=& {\Psi _x}^{ - 1}{\Psi _\theta } \\
  u &=& 1 
\end{eqnarray}
Here we have a horrible nonlinearity with term $B = \Psi _x^{ - 1}\Psi_\theta$, which is on or off depending on if the field is past the threshold. Unfortunately, this is the place where the spatial kernel parameters live. The unscented Kalman filter or a particle filter could possibly estimate C. What do you think?

@@
